Clone and graft:\\ Testing scientific applications as they are built

### 2015-01-21 turcksin-spring-mechanics/paper.tex
do not include code inside tex
instead put into subfolder and include. Update line numbers.

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
import sys
import json
import numpy

# Read the name of the input file from the command line, and read options from the file:
assert len(sys.argv) == 2, 'Please provide an input file.'
with open(sys.argv[1], 'r') as f:
    settings = json.loads(f.read())

# Then retrieve the various parameters from what we just read:
D = settings['spring constant']
L = settings['spring rest length']
x1_0 = settings['initial position'][0]
x2_0 = settings['initial position'][1]
v1_0 = settings['initial velocity'][0]
v2_0 = settings['initial velocity'][1]

m = [settings['masses'][0], settings['masses'][1]]

# describe the differential equation as a first order ODE:
y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
      v1_0[0], v1_0[1], v1_0[2], v2_0[0], v2_0[1], v2_0[2]]

def f(t, y):
    p1 = y[0:3]
    p2 = y[3:6]
    v1 = y[6:9]
    v2 = y[9:12]

    g = [0., 0., -9.81]

    dist = numpy.linalg.norm(p2-p1)
    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
    return numpy.concatenate([v1, v2, a1, a2])

# Next create an object that can integrate the ODE numerically:
start_time = 0.
end_time   = 5
import scipy.integrate
integrator = scipy.integrate.ode(f)
integrator.set_integrator('vode', rtol=1e-6)
integrator.set_initial_value(y0, start_time)

# With this, do the integration step by step, appending values to an array in each step:
t_values = [start_time]
y_values = numpy.array([y0])
while integrator.successful() and integrator.t < end_time:
    integrator.integrate(end_time, step=True)
    t_values.append(integrator.t)
    y_values = numpy.vstack((y_values, integrator.y))

# Having done so, output the number of time steps and the final positions:
print "time steps:", len(t_values)
print "final position:", y_values[len(t_values)-1,0:3], y_values[len(t_values)-1,3:6]
\end{lstlisting}

@@

When you run the program on this data, it will compute the trajectory of the
two connected balls over a number of seconds and print their final positions
to the console.
Of course, having a plot of their trajectories would be nice too, and you can
get that by adding the following lines to the bottom of the script:
\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
import matplotlib.pyplot
from mpl_toolkits.mplot3d import Axes3D
fig = matplotlib.pyplot.figure()
canvas = fig.gca(projection='3d')
canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2],
            label='body 1')
canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], 
            label='body 2')
matplotlib.pyplot.show()
\end{lstlisting}
The output this creates is shown in Fig.~\ref{fig:testcase-1}.

@@

Implementing all of this leads to the following extension of the program shown
above%
\footnote{This is the file \texttt{spring.py} in the directory corresponding to
this paper}, with the new parts in lines 21--29 and 45--46:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}

>>>>

To implement a program that can deal with this input, we first need to
describe how Python solves ODEs. Basically, it requires us to reformulate it
as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
setting $\mathbf
y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
f$ then consists of the $\mathbf v_i$ and the right hand sides of
\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:

\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
{codeinclude/spring1.py}

@@

When you run the program on this data, it will compute the trajectory of the
two connected balls over a number of seconds and print their final positions
to the console.
Of course, having a plot of their trajectories would be nice too, and you can
get that by adding the following lines to the bottom of the script:
\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
{codeinclude/plot.py}
The output this creates is shown in Fig.~\ref{fig:testcase-1}.

@@

Implementing all of this leads to the following extension of the program shown
above%
\footnote{This is the file \texttt{spring.py} in the directory corresponding to
this paper}, with the new parts in lines 19--20 and 35--36:
\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}
